From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-27 upi.tex
pi2

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the questions of when two programs are ``equivalent.'' Consider
the six programs of type $\bt \leftrightarrow \bt$ below:

@@

\noindent We neither expect, nor desire, that all these six programs
be equivalent. However each group of three programs arguably
represents ``the same'' program. Thinking extensionally, it is indeed
easy to verify this fact but we can do better. We can provide
\emph{evidence} (i.e., a program that manipulates programs) that can
constructively transform one program to an equivalent one. We show
such a level-2 program proving that $\AgdaFunction{not₃}$ is
equivalent to $\AgdaFunction{not}$:

@@

We present a small $\Pi$-based language which we will use in the formalization
in the rest of the paper. The restriction of $\Pi$ to the case of just booleans
is the following:

\[\begin{array}{rcl}
\tau &::=& \bt \\
\\
v &::=& \begin{array}[t]{lrcl}
                    & \fc &:& \bt \\
              \alt & \tc &:& \bt
               \end{array} \\
\\
c &::=& \begin{array}[t]{llcl}
              & \id &:& \tau \iso \tau \\
               \alt & \swap &:& \bt \iso \bt \\
               \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)
               \end{array} \\
\\
\alpha &::=& \begin{array}[t]{lrcl}
               & \id &:& c \isotwo c \\
            \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
            \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
            \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
            \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
            \alt & \rho &:& \swap \circ \swap \isotwo \id \\
            \alt & \assocc &:& \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
            \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
            \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
            \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
             \end{array}
\end{array}\]

>>>>

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the question of when two programs are ``equivalent.'' Consider
the six programs of type $\bt \leftrightarrow \bt$ below:

@@

\noindent We neither expect, nor desire, that all these programs of
the same type to be be equivalent. In the case above, each group of
three programs does represent ``the same'' program but the two groups
represent distinct programs. Thinking extensionally, i.e., by looking
at all possible input-output pairs, it is indeed easy to verify this
fact but we can do better. We can provide \emph{evidence} (i.e., a
program that manipulates programs) that can constructively transform
every program to an equivalent one. We show such a level-2 program
proving that $\AgdaFunction{not₃}$ is equivalent to
$\AgdaFunction{not}$. For illustration, the program for
$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
the reader to map the steps below to manipulations on the diagram that
would incrementally simplify it:

@@

Having illustrated the general flavor of the $\Pi$ family of
languages, we present in full detail a small $\Pi$-based language
which we will use in the formalization in the rest of the paper. The
language below is the restriction of $\Pi$ to the case of just one
type, the type of booleans.

\[\def\arraystretch{0.8}\begin{array}{rcl}
\tau &::=& \bt \\
\\
v &::=& \begin{array}[t]{lrcl}
                    & \fc &:& \bt \\
              \alt & \tc &:& \bt
               \end{array} \\
\\
c &::=& \begin{array}[t]{llcl}
              & \id &:& \tau \iso \tau \\
               \alt & \swap &:& \bt \iso \bt \\
               \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)
               \end{array} \\
\\
\alpha &::=& \begin{array}[t]{lrcl}
               & \id &:& c \isotwo c \\
            \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
            \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
            \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
            \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
            \alt & \rho &:& \swap \circ \swap \isotwo \id \\
            \alt & \assocc &:& \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
            \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
            \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
            \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
             \end{array}
\end{array}\]
